A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-12-02 abstract.tex
updates based on Claire's comments - up to p10

Spatial proteomics is the systematic study of protein sub-cellular
localisation. In this workflow, we describe the analysis of a typical
quantitative mass spectrometry-based spatial proteomics experiment
using the \Biocpkg{MSnbase} and \Biocpkg{pRoloc} Bioconductor package
suite. To walk the user through the computational pipeline, we use a
recently published experiment predicting protein sub-cellular
localisation in pluripotent embryonic mouse stem cells. We describe
the software infrastructure at hand, importing and processing data,
quality control, sub-cellular marker definition and visualisation. We
then demonstrate the application and interpretation of statistical
learning methods, including novelty detection using semi-supervised
learning, classification, clustering and transfer learning and
conclude the pipeline with data export. The workflow is aimed at
beginners who are familiar with proteomics in general and spatial
proteomics in particular.

>>>>

Spatial proteomics is the systematic study of protein sub-cellular
localisation. In this workflow, we describe the analysis of a typical
quantitative mass spectrometry-based spatial proteomics experiment
using the \Biocpkg{MSnbase} and \Biocpkg{pRoloc} Bioconductor package
suite. To walk the user through the computational pipeline, we use a
recently published experiment predicting protein sub-cellular
localisation in pluripotent embryonic mouse stem cells. We describe
the software infrastructure at hand, importing and processing data,
quality control, sub-cellular marker definition, visualisation and
interactive exploration. We then demonstrate the application and
interpretation of statistical learning methods, including novelty
detection using semi-supervised learning, classification, clustering
and transfer learning and conclude the pipeline with data export. The
workflow is aimed at beginners who are familiar with proteomics in
general and spatial proteomics in particular.

### 2016-12-02 bioc-workflow.Rnw
merge Lisa's latest, with overview fig
update caption plot2D/3D
updates based on Claire's comments - up to p10

Quantitative mass spectrometry based spatial proteomics involves
elaborate, expensive and time consuming experimental procedures and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets. However, data analysis
is as critical as data production for reliable and insightful
biological interpretation. Here, we walk the reader through a typical
pipeline for the analysis of such data using several Bioconductor
packages for the R statistical programming environment.

The main package to analyse protein localisation data is
\Biocpkg{pRoloc}, which offers a set of dedicated functions for the
analysis of such data. \Biocpkg{pRoloc} itself relies on
\Biocpkg{MSnbase} to manipulate and process quantitative
proteomics data. Many other packages are used by \Biocpkg{pRoloc}
for clustering, classification and visualisation.

@@

Installation of Bioconductor package is documented in detail on the
\href{http://bioconductor.org/install/#install-bioconductor-packages}{Bioconductor
  installation help page}. This procedure is also applicable to any
packages, from \href{https://cran.r-project.org/}{CRAN} as well as
GitHub. Once a package has been installed, it needs to be loaded for
its functionality to become available in the R session; this is done
with the \texttt{library} function e.g.  to load the
\Biocpkg{pRoloc} one would type \texttt{library("pRoloc")}
after installation.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/overview.png}
  \caption{Schematic overview of the pRoloc pipeline from data import, through to data processing, machine learning and data export.}
  \label{fig:overview}
\end{figure}

@@

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
\cite{hyper}. The data was generated using hyperplexed LOPIT
(hyperLOPIT), a state-of-the-art method relying on improved
sub-cellular fractionation and more accurate quantitation, leading to
more reliable classification of protein localisation across the whole
sub-cellular space. The method uses an elaborate sub-cellular
fractionation scheme, enabled by the use of Tandem Mass Tag (TMT)
\cite{Thompson:2003} 10-plex and application of the MS data
acquisition technique named synchronous precursor selection MS$^3$
(SPS-MS$^3$) \cite{McAlister:2014}, for TMT quantification with high
accuracy and precision. Three biological replicates were generated
from the E14TG2a experiment, the first was to target low density
fractions and the second and third were to emphasis separation of the
denser organelles.  The intersect of replicates 1 and 2 was treated as
a 20-plex dataset for the analysis.  As discussed in the manuscript
\cite{hyper}, it has been shown that combining replicates across from
different gradients can increase spatial resolution
\cite{Trotter:2010}. The combination of replicates resulted in 5032
proteins common in both experiments.

These, as well as many other data are directly available as properly
structured and annotated computational objects from the 
\Biocexptpkg{pRolocdata} experiment package. In this workflow, we
will start with a description of how to generate these ad hoc objects
starting from an arbitrary spreadsheets, as produced by many popular
third-party applications.

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
  capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2}. The function takes a single spreadsheet
filename as input and extracts the columns containing the quantitation
data, as identified by the argument \texttt{ecol}, to create the
expression data, while the other columns in the spreadsheet are
appended to the feature meta-data slot.  By example, in the code chunk
below we read in the \texttt{csv} spreadsheet containing the
quantitation data from the intersect of replicates 1 and 2 of the
mouse map \cite{hyper}, using the \texttt{readMSnSet2} function. The
data is as available online with the manuscript (see tab 2 of the
\texttt{xlsx} supplementary data set 1 in \cite{hyper}, which should
be exported as a text-based spreadsheet). It is also available as a
\texttt{csv} in the Bioconductor \Biocexptpkg{pRolocdata} data
package, which we make use of below.

To use the \texttt{readMSnSet2} function, as a minimum one must specify the
file path to the data and which columns of the spreadsheet contain
quantitation data. The \texttt{getEcols} function exists to help users
identify which columns of the spreadsheet contain the
quantitation data. In the last line of the code chunk below, we print
the file name (not the full path, which will vary from computer to
computer).

@@

The spreadsheet that was deposited by the authors contains two
headers, with the second header containing information about where the
quantitation data is stored.

@@

We can display the names of the second header by calling the
\texttt{getEcols} function with the argument \texttt{n = 2} (the default value is \texttt{n
= 1}), to specify that we wish to display the column names of the
second line.

@@

Below, we examine the quantitative information for first 5 proteins.
It is also possible to access specific rows and columns by naming the
proteins and TMT tag channels of interest.

@@

<<removefData>>=
fvarLabels(hl)
fvarLabels(hl)[1:3] <- c("uniprot.accession", "uniprot.id", "description")
fvarLabels(hl)[4:6] <- paste0("peptides.expt", 1:3)
fData(hl)[1:4, 1:6]
@

Note that when using the simple \texttt{readMSnSet2} procedure, the \texttt{pData}
slot which is used to store information about the samples/channels
is kept empty. It is advised to annotate the channels as
well. Below, we annotate the replicate from which the profiles
originate and the TMT tag (extracted from the sample/channel names).

@@

Throughout this workflow we refer to the different columns that
are found in the \texttt{exprs} (expression data) slot as channels (short for TMT channels).
In the frame of LOPIT and hyperLOPIT these channels consititue the
relative abundance of each protein (along the rows) in the channel of
interest. Each TMT channel originates from fractions collected from
the density gradient, or a set of pooled fractions or may be a sample
originating from an alternative preparation e.g. such as from the
chromatin enrichment performed in Christoforou et al \cite{hyper}.
Information about which gradient fractions were used for which tag
should also be stored in the sample meta-data \texttt{pData} slot.

@@

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without
affecting biological variability (i.e. the shape of the quantitative
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into a same intensity interval (typically 0 and 1) by, for
example, diving each intensity by the sum of the intensities for that
quantitative feature. This is not necessary in this example as the
intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer. However, if one wanted to do this they would
execute the \texttt{normalise} function as demonstrated in the below code
chunk.

@@

The same \texttt{normalise} function (or \texttt{normalize}, both spellings are
supported) can also be applied in the first case described above.
Different normalisation methods such as mean or median scaling,
variance stabilisation or quantile normalisation, to cite a few, can
be applied to accomodate different needs.

@@

In addition, to matching names, the content of the feature metadata
for identical feature annotations must match exactly across the data
to be combined. In particular for these data, we expect the same
proteins in each replicates to be annotated with the same UniProt
entry names and descriptions, but not with the same coverage of number
of peptides or peptide-spectrum matches (PSMs).

@@

Below, we update the replicate specific feature variable names and
remove the shared annotation.

<<fvarnames2>>=
fvarLabels(hyperLOPIT2015ms3r1)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r1)[3:5], 1)
fvarLabels(hyperLOPIT2015ms3r2)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r2)[3:5], 2)
fData(hyperLOPIT2015ms3r1) <- fData(hyperLOPIT2015ms3r1)[1:5] 
fData(hyperLOPIT2015ms3r2) <- fData(hyperLOPIT2015ms3r2)[3:5]
fvarLabels(hyperLOPIT2015ms3r1)
fvarLabels(hyperLOPIT2015ms3r2)
@

@@

More details above combining data are given in the dedicated
\textit{Combining MSnSet instances} section of the \Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf}{tutorial
  vignette}.

@@

We prefer to remove proteins that were not assayed in both replicated
experiments. This is done with the \texttt{filterNA} function that
removes features (proteins) that contain more than a certain proportion (default
is 0) missing values.

@@

\subsection*{Adding user-defined marker lists}
It is also possible for users to use their own marker list with the
\texttt{addMarkers} function. The user needs to create a named vector
of marker localisation, or a create a csv file with two columns (one
for the protein names, one for the corresponding sub-cellular marker
annotation) and pass the vector or filename respectively to the
function. As previously mentioned, the protein names of these markers
must match some (but not necessarily all) of the \texttt{MSnSet}'s
feature names. See \texttt{?addMarkers} for more details.

@@

\begin{figure}[!ht]
  \centering
<<plotmarkers, out.width="\\textwidth", fig.width=12, fig.asp=1/2>>=
par(mfrow = c(1, 2))
plot2D(hl, main = "pRolocmarkers for mouse")
addLegend(hl, cex = .6)
plot2D(hl, dims = c(1, 7), main = "Marker resolution along PC 1 and 7")
@  
  \caption{Annotated PCA plots of the \texttt{hl} dataset.}
  \label{fig:plotmarkers}
\end{figure}

@@

>>>>

Quantitative mass spectrometry based spatial proteomics involves
elaborate, expensive and time consuming experimental procedures and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets (see for example
\cite{Gatto2010} for a review, and \cite{hyper,Itzhak:2016} for recent
examples). However, data analysis is as critical as data production
for reliable and insightful biological interpretation. Here, we walk
the reader through a typical pipeline for the analysis of such data
using several Bioconductor packages for the R statistical programming
environment.

The main package to analyse protein localisation data is
\Biocpkg{pRoloc}, which offers a set of dedicated functions for the
analysis of such data. \Biocpkg{pRoloc} itself relies on
\Biocpkg{MSnbase} to manipulate and process quantitative proteomics
data. Many other packages are used by \Biocpkg{pRoloc} for clustering,
classification and visualisation. Support for interactive
visualisation is offered by the \texttt{pRolocGUI} package.

@@

Installation of Bioconductor package is documented in detail on the
\href{http://bioconductor.org/install/#install-bioconductor-packages}{Bioconductor
  installation help page}. Below, we show how to install the four main
packages used in this workflow:

<<install, eval=FALSE>>=
source("https://bioconductor.org/biocLite.R")
biocLite(c("MSnase", "pRoloc", "pRolocdata", "pRolocGUI"))
@

This procedure is also applicable to any packages, from
\href{https://cran.r-project.org/}{CRAN} as well as GitHub. Once a
package has been installed, it needs to be loaded for its
functionality to become available in the R session; this is done with
the \texttt{library} function e.g.  to load the \Biocpkg{pRoloc} one
would type \texttt{library("pRoloc")} after installation.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/overview.pdf}
  \caption{Schematic overview of the pRoloc pipeline from data import, through to data processing, machine learning and data export.}
  \label{fig:overview}
\end{figure}

@@

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
\cite{hyper}. The data was generated using hyperplexed LOPIT
(hyperLOPIT), a state-of-the-art method relying on improved
sub-cellular fractionation and more accurate quantitation, leading to
more reliable classification of protein localisation across the whole
sub-cellular space. The method uses an elaborate sub-cellular
fractionation scheme, enabled by the use of Tandem Mass Tag (TMT)
\cite{Thompson:2003} 10-plex and application of the MS data
acquisition technique named synchronous precursor selection MS$^3$
(SPS-MS$^3$) \cite{McAlister:2014}, for TMT quantification with high
accuracy and precision. Three biological replicates were generated
from the E14TG2a experiment, the first was to target low density
fractions and the second and third were to emphasis separation of the
denser organelles.  The intersect of replicates 1 and 2 was treated as
a 20-plex dataset for the analysis.  As discussed in the manuscript
\cite{hyper}, it has been shown that combining replicates from
different gradients can increase spatial resolution
\cite{Trotter:2010}. The combination of replicates resulted in 5032
proteins common in both experiments.

These, as well as many other data are directly available as properly
structured and annotated datasets from the \Biocexptpkg{pRolocdata}
experiment package. In this workflow, we will start with a description
of how to generate these ad hoc data objects starting from an
arbitrary spreadsheet, as produced by many popular third-party
applications.

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
  capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2}. The function takes a single spreadsheet
file name as input and extracts the columns containing the
quantitation data, as identified by the argument \texttt{ecol}, to
create the expression data, while the other columns in the spreadsheet
are appended to the feature meta-data slot.  By example, in the code
chunk below we read in the \texttt{csv} spreadsheet containing the
quantitation data from the intersect of replicates 1 and 2 of the
mouse map \cite{hyper}, using the \texttt{readMSnSet2} function. The
data is as available online with the manuscript (see tab 2 of the
\texttt{xlsx} supplementary data set 1 in \cite{hyper}, which should
be exported as a text-based spreadsheet). It is also available as a
\texttt{csv} in the Bioconductor \Biocexptpkg{pRolocdata} data
package, which we make use of below.

To use the \texttt{readMSnSet2} function, as a minimum one must
specify the file path to the data and which columns of the spreadsheet
contain quantitation data. In the code chunk below, we start by
identifying the file that we want to use. The \texttt{system.file}
function is used to return the path to the \texttt{extdata} directory
from the \texttt{pRolocdata} package, which is where our file of
interest resides. We then use the \texttt{dir} function to list the
content of that directory and store the path that matches the file
name of interest in the \texttt{csvfile}. Note that these two lines
are only needed here to locate a file in a package; in a more general
use case, the user would defined the \texttt{csvfile} variable
containing the file name of interest directly.

A common pitfall here is to provide only the file name, rather than
full path to the file (which is what is shown below with
\texttt{basename}; we don't print the full path, as it will vary from
computer to computer). Note that only specifying the name of the file
is sufficient when it exists in the working directory (i.e. the
directory in which R is running, which can be queried and changed with
the \texttt{getwd} and \texttt{setwd} functions respectively).

@@

Next, we need to identify which columns in the spreadsheet contain the
quantitation data. This can be done using the \texttt{getEcols}
function inside R. The spreadsheet that was deposited by the authors
contains two headers, with the second header containing information
about where the quantitation data is stored.

@@

We can display the names of the second header by calling the
\texttt{getEcols} function with the argument \texttt{n = 2} (the
default value is \texttt{n = 1}), to specify that we wish to display
the column names of the second line. We also specify the name of the
spreadsheet file (defined as \texttt{csvfile} above) and the
separator that splits cells.

@@

Below, we examine the quantitative information along the whole
gradient for first 5 proteins.  It is also possible to access specific
rows and columns by naming the proteins and TMT tag channels of
interest.

@@

<<removefData>>=
fvarLabels(hl)
fvarLabels(hl)[1:3] <- c("uniprot.accession", "uniprot.id", "description")
fvarLabels(hl)[4:6] <- paste0("peptides.expt", 1:3)
## feature vars 1, 2, and 4 to 6
fData(hl)[1:4, c(1:2, 4:6)]
@

Note that when using the simple \texttt{readMSnSet2} procedure, the
\texttt{pData} slot which is used to store information about the
samples/channels is kept empty. As illustrated below, one can use the
\texttt{\$} operator to access (or create) individual columns in the
metadata slot. It is advised to annotate the channels as well. Below,
we annotate the replicate from which the profiles originate and the
TMT tag (extracted from the sample/channel names). To do so, we use
the sample names that were assigned automatically using the
quantiation column names and remove leading \texttt{X} and trailing
\texttt{.1} using the \texttt{sub} function.

@@

Throughout this workflow we refer to the different columns that
are found in the \texttt{exprs} (expression data) slot as channels (short for TMT channels).
In the frame of LOPIT and hyperLOPIT these channels constitute the
relative abundance of each protein (along the rows) in the channel of
interest. Each TMT channel originates from fractions collected from
the density gradient, or a set of pooled fractions or may be a sample
originating from an alternative preparation e.g. such as from the
chromatin enrichment performed in Christoforou et al \cite{hyper}.
Information about which gradient fractions were used for which tag
should also be stored in the sample meta-data \texttt{pData} slot.

@@

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without affecting
biological variability (i.e. the shape of the quantitative
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into a same intensity interval (typically 0 and 1) by, for
example, dividing each intensity by the sum of the intensities for
that quantitative feature. This is not necessary in this example as
the intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer v1.4, Thermo Fisher. However, if one wanted to do
this they would execute the \texttt{normalise} function as
demonstrated in the below code chunk.

@@

The same \texttt{normalise} function (or \texttt{normalize}, both
spellings are supported) can also be applied in the first case
described above.  Different normalisation methods, such as mean or
median scaling, variance stabilisation or quantile normalisation, to
cite a few, can be applied to accomodate different needs (see
\texttt{?normalise} for available options).

@@

In addition to matching names, the content of the feature metadata for
identical feature annotations must match exactly across the data to be
combined. In particular for these data, we expect the same proteins in
each replicate to be annotated with the same UniProt entry names and
descriptions, but not with the same coverage of number of peptides or
peptide-spectrum matches (PSMs).

@@

Below, we update the replicate specific feature variable names and
remove the shared annotation. In the first line, we update only the
feature variable names 3 to 5 (by appending a \texttt{1}) and in the
second line, we apply the \texttt{updateFvarLabels} function to update
all feature variable names (by appending a \texttt{2}).

<<fvarnames2>>=
fvarLabels(hyperLOPIT2015ms3r1)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r1)[3:5], 1)
hyperLOPIT2015ms3r2 <- updateFvarLabels(hyperLOPIT2015ms3r2, "2", sep = "")
fData(hyperLOPIT2015ms3r1) <- fData(hyperLOPIT2015ms3r1)[1:5]
fData(hyperLOPIT2015ms3r2) <- fData(hyperLOPIT2015ms3r2)[3:5]
fvarLabels(hyperLOPIT2015ms3r1)
fvarLabels(hyperLOPIT2015ms3r2)
@

@@

More details about combining data are given in the dedicated
\textit{Combining MSnSet instances} section of the \Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf}{tutorial
  vignette}.

@@

We prefer to remove proteins that were not assayed in both replicated
experiments. This is done with the \texttt{filterNA} function that
removes features (proteins) that contain more than a certain
proportion (default is 0) missing values. The \textit{Processing
  information} section summarises how combining and filtering missing
values (subsetting) changed the dimensions of the data.

@@

\subsection*{Adding user-defined marker lists}
It is also possible for users to use their own marker list with the
\texttt{addMarkers} function. The user needs to create a named vector
of marker localisation, or a create a csv file with two columns (one
for the protein names, one for the corresponding sub-cellular marker
annotation) and pass the vector or file name respectively to the
function. As previously mentioned, the protein names of these markers
must match some (but not necessarily all) of the \texttt{MSnSet}'s
feature names. See \texttt{?addMarkers} for more details.

@@

\begin{figure}[!ht]
  \centering
<<plotmarkers, out.width="\\textwidth", fig.width=12, fig.asp=1/2>>=
par(mfrow = c(1, 2))
plot2D(hl, main = "pRolocmarkers for mouse")
addLegend(hl, cex = .6)
plot2D(hl, dims = c(1, 7), main = "Marker resolution along PC 1 and 7")
@  
\caption{Annotated PCA plots of the \texttt{hl} dataset, as produced
  with \texttt{plt2D} and \texttt{plot3D} (right).}
  \label{fig:plotmarkers}
\end{figure}

@@

We also recommend that users regularly update the packages as well as
the R itself. This can be done with the \texttt{biocLite} function.

<<update, eval=FALSE>>=
library("BiocInstaller")
biocLite()
@
